---
title: "Non-Euclidean space as revealed by proper distances from Mercury to Venus"
author: "Robin Hankin"
date: "2023-05-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here I consider the difference between two ways to assess the distance
between Mercury and Venus.  We use units in which the Schwarzschild
radius is unity, converting to SI when needed.  Astronomically, the
correct values are

```{r}
M <- 1.989e30 # mass of Sun in kg
G <- 6.6743e-11 # G in SI
sol <- 299792458 # speed of light
(r <- 2*M*G/sol^2) # Schwarzschild radius of Sun in m
(mercury <- 57909050e3/r) # semimajor axis
(venus <- 108208000e3/r)
```

For simplicity, I will first consider Mercury and Venus to have an
orbital radius of a notional $2\times 10^7$ and $4\times 10^7$
respectively.  The Euclidean distance between the two is just $4\times
10^7-2\times 10^7=2\times 10^7$ but the proper distance is given by

$$
\int_{r=2\times 10^7}^{4\times 10^7}\frac{dr}{\sqrt{1-1/r}}
$$

Mathematically this is easy.  Mathematica gives

```
In[27]:= Assuming[x>0,Integrate[1/Sqrt[1-1/r],{r,2*10^7,4*10^7}]]                                                                                                                     
Out[27]= -2000 (Sqrt[99999995] - Sqrt[399999990]) - ArcSinh[Sqrt[19999999]] + ArcSinh[Sqrt[39999999]]
```

Numerically we see close agreement between Mathematica and R:

```
In[28]:= p = Assuming[x>0,Integrate[1/Sqrt[1-1/r],{r,2*10^7,4*10^7}]]                        
In[29]:= (p - 2*10^7)//N
Out[29]= 0.346574
```

```{r}
(v <- integrate(function(r){1/sqrt(1-1/r)},2e7,4e7))
v$value - 2e7
```


The discrepancy is in Schwarzschild radii.

```{r}
(v$value-2e7)*r
```

This gives us the discrepancy in meters.  We see a proper distance
between the two planets of about a kilometer in excess of the
Euclidean distance calculated on the basis of circumferential radius
and flat space.  It's not that much harder to use the astronomically
accurate values instead:

```{r}
diff <- (integrate(function(r){1/sqrt(1-1/r)},mercury,venus)$value - (venus-mercury))
diff   # discrepancy in Schwarzschild radius
diff*r  # in meters
```

Asymptotically we have

$$
\int_{r=w}^{wx}\frac{dr}{\sqrt{1-1/r}}=w(x-1)+\frac{\log(x)}{2} -w^{-1}\frac{3(1-x)}{8x}+{\mathcal O}\left(w^{-2}\right)
$$
(writing $w$ for Mercury's orbital radius).  Noting that $x\sim 1$, the first two terms are about 

```{r}
x <- venus/mercury
log(x)/2
3*(1-x)/(8*mercury*x)
```

showing that the zeroth order term is sufficient here, and also close
agreement with the numerical integration using astronomical data given
above.